Convergence rate of MAP estimates \\ for the exponential family

### 2022-02-10 colt/paper_content.tex
Remove some space but still 13 pages
Update paper_content.tex

\begin{figure}[ht]
	\centering
\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial}, 
or more recently  $\chi^2$ distance \citep{kamath2015learning} or Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
and raising the fundamental problem:
\begin{equation}
\boxed{
	\textit{Find an upper bound on the expected value of}\ \ \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
}
\tag{$\star$}
\label{problem}
\end{equation}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution. \vspace{-3ex}
	}
	\label{fig:thales}
\end{figure}

@@

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
    &\leq 2L \norm{\m^* - \m} \; ,
\end{align}
so $\bregmanconj$ is Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate, but no common exponential families verify the assumption.

{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    \leq \frac{L}{2} \norm{\m^* - \m}^2 \; ,
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
% It is also possible to get upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function. The next section focuses on \textit{local} quadratic behavior, which is more realistic.

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From~\eqref{eq:bregmanTaylor}, we know that Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance of $\conj$ (in 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$ \citep[Ch.~4.1]{nesterov2003introductory}), we can quantify when this quadratic behavior kicks in. % Proofs for this subsection are in App.~\ref{app:self-concordant}.
\begin{proposition}
	\label{prop:selfConcordant}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then \vspace{-3ex} % \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.} 
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*; \m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 25d + 5 \norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

@@

\begin{table}[t]
	\caption{Summary of results for SMD
		under relative smoothness and relative strong convexity assumptions.
		Each row correspond to one analysis, and each columns answers one question.
		($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
		($\hat\nat_n$) is it for the last iterate or an average ?
		None of these analysis check all the boxes needed to address~\eqref{problem}.
	}
	\begin{center}
		\begin{tabular}{lcccc}
			%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
			\toprule
			Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
			\midrule
			%Strongly convex and bounded variance\newline
			%\citep[e.g.,][]{??}
			%&
			%$
			%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
			%\quad
			%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
			%$
			%\\
			Variance on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
			& \redxmark & \greencmark & \greencmark  & \redxmark
			\\
			Variance at $\theta^*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
			& \redxmark & \greencmark & \redxmark  & \greencmark
			\\
			Optimization gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
			& \greencmark & \redxmark & \redxmark & \greencmark
			\\
			\bottomrule
		\end{tabular}
	\end{center}
	\label{tbl:assumptions}
\end{table}

>>>>

\begin{figure}[ht]
	\centering
\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
	\vspace{-3ex}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial}, 
or more recently  $\chi^2$ distance \citep{kamath2015learning} or Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
and raising the fundamental problem:
\begin{equation}\setlength\fboxsep{2ex}
\boxed{
	\textit{Find an upper bound in expectation on}\ \ \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; \textit{(or in high probability)}.
}
\tag{$\star$}
\label{problem}
\end{equation}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{
	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
	}
	\label{fig:thales}
\end{figure}

@@

{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\m^* ; \m)  \leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m}  2L \norm{\m^* - \m} \; ,
\end{align}
so $\bregmanconj$ is Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate, but no common exponential families verify the assumption.

{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
\begin{align}
    \bregmanconj(\m^* ; \m)
    \leq \frac{L}{2} \norm{\m^* - \m}^2 \; ,
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
% It is also possible to get upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function. The next section focuses on the more realistic \textit{local} quadratic behavior.

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From~\eqref{eq:bregmanTaylor}, we know that Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance of $\conj$ (in 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$ \citep[Ch.~4.1]{nesterov2003introductory}), we can quantify when this quadratic behavior kicks in. % Proofs for this subsection are in App.~\ref{app:self-concordant}.
\begin{proposition}
	\label{prop:selfConcordant}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then % \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.} 
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*; \m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 25d + 5 \norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

@@

\begin{table}[t]
	\begin{center}
		\begin{tabular}{lcccc}
			%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
			\toprule
			Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
			\midrule
			%Strongly convex and bounded variance\newline
			%\citep[e.g.,][]{??}
			%&
			%$
			%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
			%\quad
			%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
			%$
			%\\
			Variance on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
			& \redxmark & \greencmark & \greencmark  & \redxmark
			\\
			Variance at $\theta^*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
			& \redxmark & \greencmark & \redxmark  & \greencmark
			\\
			Optimization gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
			& \greencmark & \redxmark & \redxmark & \greencmark
			\\
			\bottomrule
		\end{tabular}
	\end{center}
	\vspace{-3ex}
	\caption{Summary of results for SMD
		under relative smoothness and relative strong convexity assumptions.
		Each row correspond to one analysis, and each columns answers one question.
		($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
		($\hat\nat_n$) is it for the last iterate or an average ?
		None of these analysis check all the boxes needed to address~\eqref{problem}.
	}
	\label{tbl:assumptions}
\end{table}
